---
title: "anova_VS_Anova"
author: "Ming Yang"
output:
  html_document:
    toc: true
    toc_depth: 2
    #theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning = F,cache=F)

```

# anova VS Anova, or Type I tests VS Type-II tests

code modified from `hiercourse` course by Remko Duursma, Jeff Powell:

- http://www.hiercourse.com/
- http://www.hiercourse.com/mixedeffects.html

```{r}
litter = read.csv('masslost.csv') #dataset download from above link
dim(litter)
head(litter)

litter$plot = as.factor(litter$plot)
litter$block = as.factor(litter$block)
litter$profile = as.factor(litter$profile)
litter$herbicide = as.factor(litter$herbicide)

# there is a date column
library(lubridate)
litter$date=mdy(litter$date)
litter$date2=litter$date-ymd('2006-05-23')

# visualize the data to look for treatment effects
library(lattice)
bwplot(masslost ~ factor(date) | profile:herbicide,data=litter)

# count the data to confirm that the design is unbalanced
xtabs(~date2 + profile + herbicide,data=litter)
ftable(xtabs(~date2 + profile + herbicide,data=litter))

# fit linear model
m1fix=lm(masslost ~ date2+herbicide*profile+block,data=litter)
anova(m1fix)
m2fix=lm(masslost ~ date2+profile*herbicide+block,data=litter)
anova(m2fix)
```

Sums of squares and p-values differ for profile and herbicide across the two fits.

This is due to the order of the variables entered in the model.

As anova performs Type-I tests, each next term is tested against a model that includes all terms preceding it.

Type-II tests implemented by Anova funciton in car package solves this issue.

In such tests, each main effect is tested against a model that includes all other terms.

```{r}
library(car)
Anova(m1fix,test='F')
Anova(m2fix,test='F')
```

If the data is balanced, the sequential (Type-I) and Type-II tests would be identical.

Select 16 items under each condition to confirm above statement

```{r}
attach(litter)
new.df=as.numeric()
for(i.date in unique(date2)){
  for(i.profile in unique(profile)){
    for(i.herbicide in unique(herbicide)){
      x=litter[date2==i.date & profile==i.profile & herbicide==i.herbicide,]
      set.seed(123)
      new.df=rbind(new.df,x[sample(1:nrow(x),16,replace=F),])
    }
  }
}
dim(new.df)
ftable(xtabs(~date2 + profile + herbicide,data=new.df))

# fit linear model
m1fix=lm(masslost ~ date2+herbicide*profile+block,data=new.df)
anova(m1fix)
m2fix=lm(masslost ~ date2+profile*herbicide+block,data=new.df)
anova(m2fix)
```

```{r}
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
```

