---
title: "lnR_metabolomics: PCA analysis"
author: "Ming Yang"
output:
  html_document:
    code_folding: hide
    #css: style.css
    toc: true
    toc_depth: 2
    df_print: paged
    #theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,message = F,warning = F,
        fig.width=6,fig.height=4,cache = TRUE,
        #fig.show='hold',
        fig.align='center')

knitr::opts_knit$set(root.dir = getwd())
options(stringsAsFactors = F)
library(ggplot2)
library(gridExtra)
```

# Data description

Data dimension: 393 metabolites X 74 samples

**60 samples in total**

- 5 genotypes: wt, wt/353, 74/211, 211/19, 19/74
- 2 ages: 15d, 30d 
- 3 batches: 1, 2, 3

2 replicates per genotype per age per batch

**Note on genotypes:**

- wt is wildtype

- wt/353 is the 'Mode 2' genotype (longevity but normal growth, reproduction, pAkt/insulin sensitivity). This genotype may provide 'longevity assurance'

- 19/74, 74/211, 211/19 are 'Mode 1' transheterozygous genotypes using three mutation alleles (E19, 74, 211). These have longevity but reduced growth and reproduction along with low pAkt/insulin resistance. These genotypes may extend lifespan by reducing costs of reproduction.

**14 controls**

- QC(I): instrument quanlity control
  + QC(I)#1, QC(I)#2, etc. indicate the same instrument quality control sample analyzed repeatedly, every 10 samples. 
  + A pooled commercial human plasma sample was used as the instrument quality control to monitor the instrument stability during the analysis.

- QC(S): sample quality control
QC(S)#1, QC(S)#2, etc. indicate the same sample quality control sample analyzed repeatedly, every 10 samples. 
A pooled sample made using a small portion from each of your samples was used as the sample quality control.

**32 SILIS **

32 Spiked Stable Isotope Labeled Internal Standards 

# Data Reproducibility
CV: coefficient of variation, sd/mean

- QC(I)
  + average CV: 0.0714
  + median CV: 0.0647
- QC(S)
  + average CV: 0.0806
  + median CV: 0.0740
- SSILIS
  + the global average of SSILIS across all 60 samples was calculated
  + the average of SSILIS per sample was calculated
  + all sample's mean SSILIS is within 15% of the global average
  
# Metabolites filter

missing values distribution

```{r}
input=readRDS('mz.dat.rds')
attach(input)

## combine data by running order 
run.order=protein.abundance$`Original Sample ID`
#sum(sample.id %in% run.order) #all sample ID match
#sum(colnames(control) %in% run.order) #14 control samples' ID match
#sum(rownames(control)==rownames(dat))
#sum(control$`Current MS Compounds NA` == mz$COMPOUND) #393 mz row names match

tmp=cbind(control,dat)
df.com=tmp[,run.order]
x=lapply(df.com,as.numeric)
df=as.data.frame(Reduce(`cbind`,x))
colnames(df)=colnames(df.com)
rownames(df)=control$`Current MS Compounds NA`

## remove SSILIS
df=df[-which(is.na(control$`KEGG ID NA`)),]

## remove mz with >5% missing measures
tmp=df[,-grep('QC',colnames(df))]
x=apply(tmp,1,function(i) sum(is.na(i)))
x1=as.data.frame(table(x))
colnames(x1)=c('number.of.NA','number.of.metabolites')
x1
df=df[x/ncol(tmp)<0.05,]
#dim(df)
```

remove 191 metabolites with missing rate >5%

**170 metabolites retained.**

Note:
The almost binary pattern is frequently observed in metabolomics, a metabolites is either detected in almost all samples or none of those samples.
The missingness pattern in global metabolomics could be more complicated, different metabolites have different levels of missingness across samples. A commonly uesd approach in analyzing global metabolomics is to remove those metabolites with missingness larger than 5%, then imputate data for the remaining missing values.

# plot mz by running order

Only show two as exmples.

```{r fig.width=12}
group=rep('sample',ncol(df))
group[grep('QC\\(I\\)',colnames(df))]='QC1';
group[grep('QC\\(S\\)',colnames(df))]='QC2';
plots=lapply(1:nrow(df),function(i){
    mz.name=rownames(df)[i]
    tmp=data.frame('sample.id'=colnames(df),'mz'=as.numeric(df[i,]))
    tmp$group=group;
    tmp$sample.id=factor(tmp$sample.id,levels=tmp$sample.id)
    ggplot(tmp,aes(x=sample.id,y=mz,col=group))+geom_point(size=2)+theme_bw(base_size = 9)+
      theme(axis.text.x = element_text(angle = 45,hjust=1))+ggtitle(mz.name)
  })
  plots[[1]]
  plots[[2]]
  if(F){
    (n=ceiling(length(plots) / 4)) #4 plots one page
    for(i in 1:n){
      start=i*4-3;
      end=i*4;
      if(i*4>=length(plots)){end=length(plots)}
      do.call(grid.arrange,c(plots[start:end],ncol=1))
    }
  }
```

# Data processing
- log-transformation
- Normalize by sample

```{r message=F,results='hide'}
dat=readRDS('./mz.filter.dat.rds')
df.mat=dat$edata;
pheno=dat$pheno

## log transform
summary(as.numeric(unlist(df.mat))) #no zero, log directly
df.mat.log=log(df.mat)

## normalize by sample
df.scaled=scale(df.mat.log,center=T,scale=T)
apply(df.scaled,2,mean) #confirm
apply(df.scaled,2,sd) #confirm

```

# PCA analysis

## before batch correction
```{r results='hide',fig.width=9,fig.height=6}
## PCA, view by batch, by genotype, by age
pca.out=prcomp(t(df.scaled)) #feature as column

dim(pca.out$rotation) #loading of orginal features of PC
colnames(pca.out$rotation)
dim(pca.out$x) #new coordinates for each sample
colnames(pca.out$x)
rownames(pca.out$x) #MTxx samples

pc.prop=pca.out$sdev/sum(pca.out$sdev)
sum(pc.prop)
df.pc=data.frame(PC=paste0('PC',1:length(pc.prop)),explained.var.prop=pc.prop/sum(pc.prop))
df.pc$PC=factor(df.pc$PC,levels=df.pc$PC)
pc.plot=ggplot(df.pc,aes(x=PC,y=explained.var.prop*100))+geom_bar(stat='identity')+theme_bw(base_size = 12)+
  theme(axis.text.x = element_text(angle = 45,hjust=1))
pc.plot
sum(df.pc$explained.var.prop[1:2]) #20.17% 

new.coord=as.data.frame(pca.out$x[,1:3])
sum(pheno$sample==rownames(new.coord))
new.coord$age=pheno$age;new.coord$genotype=pheno$genotype;
new.coord$batch=pheno$batch

p=ggplot(new.coord,aes(x=PC1,y=PC2))+geom_point()+theme_bw(base_size = 14)+
  xlab(paste0('PC1 (',round(df.pc[1,2],4)*100,'%)'))+
  ylab(paste0('PC2 (',round(df.pc[2,2],4)*100,'%)'))
p+geom_point(aes(col=factor(batch)),size=4)
p+geom_point(aes(col=age),size=4)
p+geom_point(aes(col=genotype),size=4)
p+geom_point(aes(col=genotype,shape=age),size=4)
```

Use `pairs()` function to view potential batch effect in the top 6 PCs.

```{r fig.width=14,fig.height=12}
#pairs(pca.out$x[,1:6]) #color by genotype, age, batch, run.order
pairs(pca.out$x[,1:6],col=pheno$batch,pch=16,main='by batch')
#pairs(pca.out$x[,1:6],col=factor(pheno$genotype),pch=16,main='by genotype')
#pairs(pca.out$x[,1:6],col=factor(pheno$age),pch=16,main='by age')
```

`PC1 vs PC2` and `PC1 vs PC3` shows batch effect.

## estimate batch effect using ANOVA for each metabolite

To get a quantitative measure of batch effect,
for each metabolite, I fit a linear model,
`lm(mz~batch+age+genotype,data=tmp)`,
then use ANOVA to estiamte the variance explained by `batch`.

use one metabolite as an example
```{r}
## estimate batch effect using ANOVA
mz=df.scaled[1,]
tmp=cbind(mz,pheno)
fit1=lm(mz~batch+age+genotype,data=tmp)
summary(fit1)
x=anova(fit1)
x
```

`Sum Sq by batch / sum(Sum Sq)`

```{r}
x$`Sum Sq`[1]/sum(x$`Sum Sq`)
```

apply this to all metabolites, have a look at the distribution of variance explained by batch:
```{r}
batch.var=sapply(1:nrow(df.scaled),function(i){
  mz=df.scaled[i,]
  tmp=cbind(mz,pheno)
  fit1=lm(mz~batch+age+genotype,data=tmp)
  summary(fit1)
  x=anova(fit1)
  x$`Sum Sq`[1]/sum(x$`Sum Sq`)
})
summary(batch.var)
hist(batch.var,breaks=20)
```

## after batch correction
use the `combat` function in R package `sva` to correct for batch effect.
```{r results='hide'}
library(sva)

batch=pheno$batch
modcombat = model.matrix(~1, data=pheno)
combat_edata = ComBat(dat=df.scaled, batch=batch, 
                      mod=modcombat, par.prior=TRUE, prior.plots=FALSE)

dim(combat_edata) #170mz X 60samples
df.scaled[1:3,1:3]
combat_edata[1:3,1:3]
```

```{r results='hide',fig.width=14,fig.height=12}
## perform PCA on batch corrected data
pca.out2=prcomp(t(combat_edata)) #feature as column

pairs(pca.out2$x[,1:6],col=pheno$batch,pch=16,main='by batch')
pairs(pca.out2$x[,1:6],col=factor(pheno$genotype),pch=16,main='by genotype')
pairs(pca.out2$x[,1:6],col=factor(pheno$age),pch=16,main='by age')

#dim(pca.out2$rotation) #loading of orginal features of PC
#dim(pca.out2$x) #new coordinates for each sample
#colnames(pca.out2$x) #PC
#rownames(pca.out2$x) #MTxx samples
```

```{r results='hide',fig.width=9,fig.height=6}
pc.prop=pca.out2$sdev/sum(pca.out2$sdev)
#sum(pc.prop)
df.pc=data.frame(PC=paste0('PC',1:length(pc.prop)),explained.var.prop=pc.prop/sum(pc.prop))
df.pc$PC=factor(df.pc$PC,levels=df.pc$PC)
pc.plot=ggplot(df.pc,aes(x=PC,y=explained.var.prop*100))+geom_bar(stat='identity')+theme_bw()+
  theme(axis.text.x = element_text(angle = 45,hjust=1))
pc.plot
sum(df.pc$explained.var.prop[1:2]) #20.67% 

new.coord=as.data.frame(pca.out2$x[,1:3])
sum(pheno$sample==rownames(new.coord))
new.coord$age=pheno$age;new.coord$genotype=pheno$genotype;
new.coord$batch=pheno$batch
p=ggplot(new.coord,aes(x=PC1,y=PC2))+geom_point()+theme_bw(base_size = 14)+
  xlab(paste0('PC1 (',round(df.pc[1,2],4)*100,'%)'))+
  ylab(paste0('PC2 (',round(df.pc[2,2],4)*100,'%)'))
p+geom_point(aes(col=factor(batch)),size=4)
p+geom_point(aes(col=age),size=4)
p+geom_point(aes(col=genotype),size=4)
p+geom_point(aes(col=genotype,shape=age),size=4)
```

## contributions of metabolites to PC1 and PC2

An overview of contributions (loading values) of all 170 metabolites to PC1~3.

```{r fig.width=9,fig.height=16}
## check mz loading for PC1, PC2, PC3.
#dim(pca.out2$rotation) #170x60,loading of orginal features of PC
#pca.out2$rotation[,1] #loading of mz to PC1
pheatmap::pheatmap(pca.out2$rotation[,1:3],cluster_cols=F,fontsize_row=4)
```

An overview of absolute contributions (absolute loading values) of all 170 metabolites to PC1~3.
```{r fig.width=9,fig.height=16}
pheatmap::pheatmap(abs(pca.out2$rotation[,1:3]),cluster_cols=F,fontsize_row=4)
```

I ranked metabolites using their **absolute contribution (loading) values** to PC1~3.

Below are the top 10 metabolites to PC1, top 10 metabolites to PC2, and top 10 metabolites to PC3.

```{r}
x1=order(abs(pca.out2$rotation[,1]),decreasing = T)
x1=data.frame(pca.out2$rotation[x1[1:10],1])
colnames(x1)='contribution to PC1'
x1

x2=order(abs(pca.out2$rotation[,2]),decreasing = T)
x2=data.frame(pca.out2$rotation[x2[1:10],1])
colnames(x2)='contribution to PC2'
x2

x3=order(abs(pca.out2$rotation[,3]),decreasing = T)
x3=data.frame(pca.out2$rotation[x3[1:10],1])
colnames(x3)='contribution to PC3'
x3
```