---
title: "lnR_metabolomics: univariate analysis"
author: "Ming Yang"
output:
  html_document:
    code_folding: hide
    #css: style.css
    toc: true
    toc_depth: 2
    df_print: paged
    #theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,message = F,warning = F,
        fig.width=16,fig.height=4,cache = TRUE,
        #fig.show='hold',
        fig.align='center')

knitr::opts_knit$set(root.dir = getwd())
options(stringsAsFactors = F)
library(ggplot2)
library(gridExtra)
library(car)
```

# Data description and visualization

Data dimension: 170 metabolites X 60 samples

**60 samples in total**

- 5 genotypes: wt, wt/353, 74/211, 211/19, 19/74
- 2 ages: 15d, 30d 
- 3 batches: 1, 2, 3

2 replicates per genotype per age per batch


**I combine 5 genotypes into 3 modes**

 - wt: wildtype
 - mode.1: 74/211, 211/19, 19/74. These genotypes may extend lifespan by reducing costs of reproduction.
 - mode.2: This genotype may provide 'longevity assurance'
 
Plot two metabolite's abundance among different modes and ages as examples
```{r fig.width=16}
dat=readRDS('./mz.filter.dat.rds')
df.mat=dat$edata;
pheno=dat$pheno

## log transform
#summary(as.numeric(unlist(df.mat))) #no zero, log directly
df.mat.log=log(df.mat)

## normalize by sample
df.scaled=scale(df.mat.log,center=T,scale=T)
#apply(df.scaled,2,mean) #confirm
#apply(df.scaled,2,sd) #confirm

## genotype explain
#wt is wildtype
#wt/353 is the ‘Mode 2’, no trade-off
#the other three are the 'Mode 1', extend lifespan by reducing costs of reproduction.
pheno$mode='mode.1';
pheno[pheno$genotype=='wt/353',]$mode='mode.2'
pheno[pheno$genotype=='wt',]$mode='wt'
#table(pheno$mode)
#mode.1 mode.2     wt 
#36     12     12 

group.levels=c('wt 15d', 'wt 30d','mode.1 15d','mode.1 30d' ,'mode.2 15d', 'mode.2 30d')
# select two metabolites
pick.mz=c('Sarcosine','Reduced Glutathione');
df.tmp=df.scaled[pick.mz,]
plots=lapply(1:nrow(df.tmp),function(i){
    mz=df.tmp[i,]
    mz.name=rownames(df.tmp)[i]
    tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode),batch=factor(pheno$batch));
    tmp$group=factor(paste(tmp$mode,tmp$age))
    tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
    tmp$group=factor(tmp$group,levels=group.levels)
    ggplot(tmp,aes(x=mode,y=mz,col=age,group=group,shape=batch))+
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(size=3,position = position_jitterdodge())+
      theme_bw(base_size = 14)+ggtitle(mz.name)
  })
  do.call(grid.arrange,c(plots,ncol=2))

  plots=lapply(1:nrow(df.tmp),function(i){
    mz=df.tmp[i,]
    mz.name=rownames(df.tmp)[i]
    tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode),batch=factor(pheno$batch));
    tmp$group=factor(paste(tmp$mode,tmp$age))
    tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
    tmp$group=factor(tmp$group,levels=group.levels)
   ggplot(tmp,aes(x=age,y=mz,col=mode,group=group,shape=batch))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
  })
  do.call(grid.arrange,c(plots,ncol=2))
  
```

Look at these two plots, it appears that both `mode` and `age` have a strong affect on `Sarcosine`, and only `mode` affects `Reduced Glutathione`.

In the following analysis, I performed linear regression analysis on each metabolite to identify those metabolites with a significant `age` or `mode` or `age:mode interaction` effect.

# linear regression on each metabolite with batch effect removed using `combat` function from `sva` R package

- I performed linear regression on each metabolite using:
**metabolite.abundance ~ age + mode + age:mode + error **

- then used `Anova` to test the significance of each term, i.e., age, mode, age:mode interaction term.

- The significance, or the P value, of each term for each metabolite was adjusted using `p.adjust` function in R with the `BH` method, referred to as `p.adj`.

```{r}
df.res=readRDS('mz.filter.combat.rds')
### use genotype and age as predictor
out=lapply(1:nrow(df.res),function(i){
  mz=df.res[i,]
  mz.name=rownames(df.res)[i] # fit model using df.res with age and mode as predictor, including interaction term
  tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode));
  tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
  x=lm(mz~age*mode,tmp )
  x
})

# use Anova to get p values for each factor of each mz
#Anova(out[[1]])
sig.out=lapply(out,function(x){ #make a table
  x1=Anova(x)
  #x1=summary(x) #this gives you coef, coef.p.value for each term
  pvalue=x1$`Pr(>F)`
  #terms=rownames(x1)
  pvalue
})

df.sig.out=Reduce(`rbind`,sig.out)
x1=Anova(out[[1]])
colnames(df.sig.out)=rownames(x1)
rownames(df.sig.out)=rownames(df.res)
df.sig.out=df.sig.out[,-4] #remove Residuals column
#df.sig.out.adj=data.frame(matrix(p.adjust(df.sig.out,method = 'BH'),ncol=ncol(df.sig.out)))
df.sig.out.adj=data.frame(apply(df.sig.out,2,function(i) p.adjust(i,method='BH')))
rownames(df.sig.out.adj)=rownames(df.sig.out)
colnames(df.sig.out.adj)=colnames(df.sig.out)
```

After regression, for each metabolite, there are 3 p.adj values, one for a predictor term:
```{r}
df.sig.out.adj[c(rownames(df.sig.out.adj)[1:3],pick.mz),]
```

Have a look at the P.adj distributions for these three temrs:
```{r}
par(mfrow=c(1,3))
hist(df.sig.out.adj$age,main='p.adj for Age')
hist(df.sig.out.adj$mode,main='p.adj for Mode')
hist(df.sig.out.adj$`age:mode`,main='p.adj for Age:Mode')
```

For these 170 metabolites, after fitting this linear model, the number of metabolites that were significantly affected by terms at a p.adj<0.05 cutoff:

```{r}
#nrow(df.sig.out.adj) #170
apply(df.sig.out.adj,2,function(i){sum(i<0.05)})
#age     mode age:mode 
#91      142       17 
#rownames(df.sig.out.adj[df.sig.out.adj$mode<0.05,])
#df.sig.out.adj['Sarcosine',]
```


# Groups of metabolites that are affected by age, mode or their interactions.

## metabolites affected by `age` alone


```{r fig.width=16}
#Select some metablites to visualize the model fitting (plot predictions)
#visreg: Visualization of regression functions
library(visreg)
#visreg(out[[1]],'genotype',by='age',overlay=T)

# some metabolites with only age effect
x1=df.sig.out.adj$age;x2=df.sig.out.adj$mode;x3=df.sig.out.adj$`age:mod`;
i=which(x1<0.05 & x2>0.05 & x3>0.05)
cat(length(i),'metabolites were only affected age\n')
```

Select one metabolite and plot its original abundance values (batch effect corrected):

```{r}
pick=i[1]
mz.name=rownames(df.sig.out.adj[pick,])

mz=df.res[pick,]
tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode),batch=factor(pheno$batch));
tmp$group=factor(paste(tmp$mode,tmp$age))
tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
tmp$group=factor(tmp$group,levels=group.levels)

p1=ggplot(tmp,aes(x=mode,y=mz,col=age,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
p2=ggplot(tmp,aes(x=age,y=mz,col=mode,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
do.call(grid.arrange,c(list(p1,p2),ncol=2))
```

Plot model predicted values:
```{r}
# Model predictions as a function of age, for the three modes
p1=visreg(out[[pick]],'age',by='mode',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
p2=visreg(out[[pick]],'mode',by='age',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
do.call(grid.arrange,c(list(p1,p2),ncol=2))
```

## metabolites affected by `mode` alone



```{r}
# some metabolites with only mode effect
x1=df.sig.out.adj$age;x2=df.sig.out.adj$mode;x3=df.sig.out.adj$`age:mod`;
i=which(x1>0.05 & x2<0.05 & x3>0.05)
cat(length(i),'metabolites were only affected mode\n')
```


Select one metabolite. Plot original metabolite abundance values (batch effect corrected) and model predicted values:
```{r}
pick=i[1]
mz.name=rownames(df.sig.out.adj[pick,])

mz=df.res[pick,]
tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode),batch=factor(pheno$batch));
tmp$group=factor(paste(tmp$mode,tmp$age))
tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
tmp$group=factor(tmp$group,levels=group.levels)

p1=ggplot(tmp,aes(x=mode,y=mz,col=age,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
p2=ggplot(tmp,aes(x=age,y=mz,col=mode,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
do.call(grid.arrange,c(list(p1,p2),ncol=2))

p1=visreg(out[[pick]],'age',by='mode',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
p2=visreg(out[[pick]],'mode',by='age',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
do.call(grid.arrange,c(list(p1,p2),ncol=2))
```

## metabolites affected by both `age` and `mode`, but no interaction



```{r}
# some metabolites with both age and mode effect, no interaction
x1=df.sig.out.adj$age;x2=df.sig.out.adj$mode;x3=df.sig.out.adj$`age:mod`;
i=which(x1<0.05 & x2<0.05 & x3>0.05)

cat(length(i),'metabolites were affected by both age and mode, but no interaction\n')
```

Select one metabolite. Plot original metabolite abundance values (batch effect corrected) and model predicted values:
```{r}
pick=i[1]
mz.name=rownames(df.sig.out.adj[pick,])

mz=df.res[pick,]
tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode),batch=factor(pheno$batch));
tmp$group=factor(paste(tmp$mode,tmp$age))
tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
tmp$group=factor(tmp$group,levels=group.levels)

p1=ggplot(tmp,aes(x=mode,y=mz,col=age,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
p2=ggplot(tmp,aes(x=age,y=mz,col=mode,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
do.call(grid.arrange,c(list(p1,p2),ncol=2))

p1=visreg(out[[pick]],'age',by='mode',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
p2=visreg(out[[pick]],'mode',by='age',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
do.call(grid.arrange,c(list(p1,p2),ncol=2))
```

## metabolites affected by both `age` and `mode`, and their interaction


```{r}
# some metabolites with both age and mode effect, yes interaction
x1=df.sig.out.adj$age;x2=df.sig.out.adj$mode;x3=df.sig.out.adj$`age:mod`;
i=which(x1<0.05 & x2<0.05 & x3<0.05)
cat(length(i),'metabolites were affected by both age and mode, and their interaction\n')
```

Select one metabolite. Plot original metabolite abundance values (batch effect corrected) and model predicted values:

```{r}
pick=i[1]
mz.name=rownames(df.sig.out.adj[pick,])

mz=df.res[pick,]
tmp=data.frame(mz=as.numeric(mz),age=factor(pheno$age),mode=factor(pheno$mode),batch=factor(pheno$batch));
tmp$group=factor(paste(tmp$mode,tmp$age))
tmp$mode=factor(tmp$mode,levels=c('wt','mode.1','mode.2'))
tmp$group=factor(tmp$group,levels=group.levels)


p1=ggplot(tmp,aes(x=mode,y=mz,col=age,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
p2=ggplot(tmp,aes(x=age,y=mz,col=mode,group=group))+
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(size=3,position = position_jitterdodge())+
    theme_bw(base_size = 14)+ggtitle(mz.name)
do.call(grid.arrange,c(list(p1,p2),ncol=2))

p1=visreg(out[[pick]],'age',by='mode',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
p2=visreg(out[[pick]],'mode',by='age',overlay=F,main=paste0(mz.name,'\n\n'),gg=T)+theme_bw()
do.call(grid.arrange,c(list(p1,p2),ncol=2))
```
