---
title: "fly TF-target database summary"
author: "Ming Yang"
output:
  html_document:
    toc: true
    toc_depth: 2
    #theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=F,warning = F)
```

## load TF-target network data

I downloaded raw data as [Supp S1.zip](https://genome.cshlp.org/content/suppl/2012/03/23/gr.127191.111.DC1/S1_integrative_regulatory_networks.zip) from the original publication and saved to local folder `2012_data`.

Original publication : Marbach, Daniel, et al. "Predictive regulatory models in Drosophila melanogaster by integrative inference of transcriptional networks." Genome research 22.7 (2012): 1334-1349.

```{r}
df=read.table("./2012_data/S1_integrative_regulatory_networks/flynet_supervised_0.6.txt",
              as.is=T);
head(df);dim(df)
str(df)
```

## gene ID mapping: change FBgn id to gene symbol
```{r}
library(AnnotationDbi);
library(org.Dm.eg.db,verbose=F,quietly=T)
package.version("org.Dm.eg.db")
x1<-AnnotationDbi::select(org.Dm.eg.db,
          keys=df[,1],keytype="FLYBASE",c("SYMBOL","GENENAME"))
dim(x1);dim(df)
# how many genes don't have a mapped gene symbol
unique(x1[is.na(x1$SYMBOL),1])
```

## Deal with fly base secondary ID
```{r}
#check one gene ("FBgn0000054") in more detail
#googled online, FBgn0000054 symbol is Adf1
select(org.Dm.eg.db,keys='Adf1',keytype='SYMBOL',c("FLYBASE","GENENAME"))
#check another gene('FBgn0033310')
#googled online, FBgn0033310 symbol is dar1
select(org.Dm.eg.db,keys='dar1',keytype='SYMBOL',c("FLYBASE","GENENAME"))
```

As shown (online)[!https://wiki.flybase.org/wiki/FlyBase:ID_Validator], those IDs are called 'secondary ID' which is not used any more, but could be mapped to one current gene ID.

```{r eval=F}
unmap.genes<-unique(x1[is.na(x1$SYMBOL),1])
browseURL(url='https://flybase.org/convert/id')
paste(unmap.genes,collapse =' ')
```

You can paste those gene names in the query box in FlyBase.
As it is kinda messy, instead, I just throw those genes away.

## use mapped gene symbols to generate database
```{r}
x1<-AnnotationDbi::select(org.Dm.eg.db,
          keys=df[,1],keytype="FLYBASE",c("SYMBOL","GENENAME"))
dim(x1)
length(unique(x1[is.na(x1$SYMBOL),1]))

x2<-AnnotationDbi::select(org.Dm.eg.db,
       keys=df[,2],keytype="FLYBASE",c("SYMBOL","GENENAME"))
dim(x2)
length(unique(x2[is.na(x2$SYMBOL),1]))

df1<-data.frame(x1$SYMBOL,x2$SYMBOL,df$V3,stringsAsFactors = F)
df2<-df1[!is.na(df1[,1]) & !is.na(df1[,2]),]
dim(df1);#310634
dim(df2);#272822
```

## remove self loop (from and to is the same gene)
```{r}
## remove redundancy
colnames(df2)=c('from','to','link');
sum(df2$from==df2$to)

## check for edge redundancy
edge=paste(df2[,1],df2[,2])
(dups=table(edge)[table(edge)>1]) #no redundancy
head(df2)
saveRDS(df2,'./processed_flynet_supervised_0.6.rds')
```

## characterize TF-target database
```{r}
df=readRDS('./processed_flynet_supervised_0.6.rds')
tf=unique(df$from);
target=unique(df$to);

# pure TF, TF&target, pure target
library(RVenn)
library(ggplot2)
my.sets=list(tf=tf,target=target)
out=Venn(my.sets)
ggvenn(out, slice = c(1, 2))

summary(as.numeric(table(df$from)))
hist(table(df$from),main='kout')

summary(as.numeric(table(df$to)))
hist(table(df$to),main='kin')
```

```{r}
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
```

